Revealing Benefits, Costs, and Tradeoffs of Spatial Management in Marine Systems Arrising from Coupled Economic and Ecological Dynamics

### 2023-04-03 03_marlin_paper.qmd
fun with math notation latex issues

---
format:
  pdf:
    toc: false
    fig-pos: 'H'
    mainfont: Times New Roman
    sansfont: Times New Roman
    fontsize: 12pt
  html: default
  docx: default
execute:
  echo: false
  message: false
  warning: false
  cache: false
csl: chicago-author-date-16th-edition.csl
bibliography: references.bib
params:
  run_name: ["v1.0"]
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
---

@@

We provide a brief overview of the the general CTMC method here, see
@thorson2021a for a detailed description. Under this framework, movement
from each patch to each other patch in the system in a given timestep
*t* for life stage *a* of species *s* is defined by a movement matrix
$\boldsymbol{M}_{t,s,a}$. $\boldsymbol{M}_{t,s,a}$ is calculated as the
matrix exponent of a function of the diffusion $\boldsymbol{D}$ and
taxis $\boldsymbol{\tau}$ matrices scaled by the width of the time step
$\Delta_{t}$ and the length of the edge of each patch $\Delta_d$ that
the model is running under. So for example, if we are running a model on
a monthly timestep, and we define the units of time in years, then
$\Delta_{t} = 1/12$. Similarly, if we are running a model where one
patch equals 100km^2^, then $\Delta_d = 10km$. This parameterization
allows users to set the effective area of the spatial domain through two
avenues; the number of patches, which effectively scales the resolution
of the model, and the area of each patch, which scales the spatial
extent of the simulation.

$$
\boldsymbol{M}_{t,s,a} = e^{\Delta_{t}(\boldsymbol{D}_{t,s,a} + \boldsymbol{\tau}_{t,s,a})} 
$$ {#eq-movement}

The individual components (*m*) of the movement matrix are filled based
on an adjacency matrix, which defines whether two patches are both
adjacent and water (as opposed to land or another physical barrier), a
diffusion rate $D$ defined in units of area of a patch per unit of time,
and a habitat preference function *h* in units of length of a side of a
patch per unit time. For example, if we are defining the time units as
years and the distance units as kilometers, for a tuna $D$ might be
1,000 $\frac{KM^2}{Year}$. We then use parameters $\Delta_{t}$ and
$\Delta_{d}$ parameters to translate the diffusion rate $D$ to match the
time step and patch size used in a simulation. This "scale free"
parameterization means that appropriate value of $D$ can be identified
for a species and then set, regardless of the time step or patch size
used in the simulation model itself. The taxis component of the movement
process is a function of the difference in habitat quality *h*. The
habitat preference function itself can take any form the user wishes.
Exponentiating the difference in the habitat preference function between
patches turns the taxis matrix into a multiplier of the diffusion rate
*D*. As such, when creating habitat layers for simulation, users can
tune the scale of the habitat gradient function to result in realistic
multipliers of the diffusion rate. This parameterization ensure that the
off-diagonal elements of the movement matrix $\boldsymbol{M}_{t,s,a}$
are all non-negative, a requirement of the CTMC method.

$$ 
m_{p1,p2,t,s,a} = \begin{cases}
      = \frac{\Delta_{t}}{\Delta_{d}^2}De^{\frac{\Delta_t(h(p2,t,s,a) - h(p1,t,s,a))}{\Delta_d}} & \text{if p2 and p1 are adjacent}\\
     = -\sum_{p' \neq p1} m_{p1,p2,t,s,a} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
\end{cases}
$$ {#eq-diffusion}

@@

While this CTMC approach to movement simulation is to date not commonly
seen in the marine modeling literature, it has numerous advantages that
warrant its seeming complexity. First, when only diffusion is present,
the model will generalize to the familiar dynamics of a Gaussian
dispersal kernel. However, the time component *t* in @eq-movement allows
the movement model to be run on arbitrarily fine time steps. Second, the
taxis model allow for more active habitat choices by species, allowing
us to simulate preferences of species in space and time efficiently.
Lastly, this form has the advantage that the parameters are directly
estimable from real data. So, if provided with for example spatial
abundance data and a tagging study, users can estimate the diffusion and
taxis movement parameters in the same manner as @thorson2021a, and then
pass their estimated parameters to `marlin` for simulation (so long as
the estimating method uses the same functional form as the movement
model in `marlin`).

@@

$$
  n_{t,p,s,a = 1} = \left(\frac{0.8{\times}r0_{p,s}\times{h_s}\times{ssb_{t-1,p,s}}}{0.2\times{ssb0_{p,s}}\times(1 - h_s)+(h_s - 0.2)\times{ssb_{t-1,p,s}}}\right)\times \boldsymbol{d^l_s} \times \epsilon_{t,s}
$$ {#eq-dd3}

@@

$$
  larv_{t,p,s} = ssb_{t-1,p,s} \times\boldsymbol{d^l_s}
$$ {#eq-larvmove}

@@

<!-- obj_p = \Phi \times scale(\sum_f^F \boldsymbol{\phi_{f,p}}) + (1 - \Phi)  scale(\boldsymbol{\sum_s^S\frac{ssb_s}{ssb0_{s,p}}}) -->

>>>>

---
format:
  pdf:
    toc: false
    fig-pos: 'H'
    mainfont: Times New Roman
    sansfont: Times New Roman
    fontsize: 12pt
  html: default
  docx: default
execute:
  echo: false
  message: false
  warning: false
  cache: false
csl: chicago-author-date-16th-edition.csl
bibliography: references.bib
params:
  run_name: ["v1.0"]
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
  - \usepackage{bm}
---

@@

We provide a brief overview of the the general CTMC method here (see
@thorson2021a for a detailed description). Under this framework, movement of individuals
from each patch to each other patch in the system in a given timestep
*t* for life stage *a* of species *s* is defined by a movement matrix
$\pmb{M}_{t,s,a}$. $\pmb{M}_{t,s,a}$ is calculated as a function of diffusion $\pmb{D}$ and
taxis $\pmb{\tau}$ matrices scaled by the width of the time step
$\Delta_{t}$ and the length of the edge of each patch $\Delta_d$ that
the model is running under. This parameterization
allows users to set the effective area of the spatial domain through two
avenues; the number of patches, which effectively scales the resolution
of the model, and the area of each patch, which scales the spatial
extent of the simulation.

The individual components (*M*) of the movement matrix ($\pmb{M}$) are filled based
on an adjacency matrix, which defines whether two patches are both
adjacent and water (as opposed to land or another physical barrier), a
diffusion rate $D$ defined in units of area of a patch per unit of time,
and a habitat preference function *H* in units of length of a side of a
patch per unit time. For example, if we are defining the time units as
years and the distance units as kilometers, for a tuna $D$ might be
1,000 $\frac{KM^2}{Year}$. We then use parameters $\Delta_{t}$ and
$\Delta_{d}$ parameters to translate the diffusion rate $D$ to match the
time step and patch size used in a simulation. For example, if we were to run a model on
a monthly timestep given time units of years, then
$\Delta_{t} = 1/12 years$. If one square patch in the simulation has an area of 100km^2^, then $\Delta_d = 10KM$. This "scale free"
parameterization means that appropriate value of $D$ can be identified
for a species and then set, regardless of the time step or patch size
used in the simulation model itself. The taxis component of the movement
process is a function of the difference in habitat quality *H*. The
habitat preference function itself can take any form the user wishes.
Exponentiating the difference in the habitat preference function between
patches turns the taxis matrix into a multiplier of the diffusion rate
*D*. As such, when creating habitat layers for simulation, users can
tune the scale of the habitat gradient function to result in realistic
multipliers of the diffusion rate. This parameterization ensure that the
off-diagonal elements of the movement matrix $\pmb{M}_{t,s,a}$
are all non-negative, a requirement of the CTMC method.

$$ 
M_{p1,p2,t,s,a} = \begin{cases}
      = \frac{\Delta_{t}}{\Delta_{d}^2}De^{\frac{\Delta_t(H(p2,t,s,a) - H(p1,t,s,a))}{\Delta_d}} & \text{if p2 and p1 are adjacent}\\
     = -\sum_{p' \neq p1} M_{p1,p2,t,s,a} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
\end{cases}
$$ {#eq-diffusion}

@@

The movement of individuals across patches is then calculated by matrix multiplication of the pre-movement vector of the number of individuals ($\pmb{n}$) of species *s* at age *a* in time step *t* across all patches *p* times the matrix exponential of the movement matrix $\pmb{M}$

$$
\pmb{n}_{t+1,s,a} = \pmb{n}_{t,s,a}e^{\pmb{M}_{t,s,a}}
$$ {#eq-movement}

While this CTMC approach to movement simulation is to date not commonly
seen in the marine modeling literature, it has numerous advantages that
warrant its seeming complexity. First, the parameters of the model have interpretable biological meaning (e.g. the diffusion rate *D*). Second, when only diffusion is present,
the model will generalize to the familiar dynamics of a Gaussian
dispersal kernel at whatever spatial and temporal resolution the simulation is set to. Third, the
taxis model allow for clearly parameterized active habitat choices by species, allowing
us to simulate preferences of species in space and time efficiently.
Lastly, the CTMC form has the advantage that its parameters are directly
estimable from real data. So, if provided with for example spatial
abundance data and a tagging study, users can estimate the diffusion and
taxis movement parameters in the same manner as @thorson2021a, and then
pass their estimated parameters to `marlin` for simulation (so long as
the estimating method uses the same functional form as the movement
model in `marlin`).

@@

$$
  n_{t,p,s,a = 1} = \left(\frac{0.8{\times}r0_{p,s}\times{h_s}\times{ssb_{t-1,p,s}}}{0.2\times{ssb0_{p,s}}\times(1 - h_s)+(h_s - 0.2)\times{ssb_{t-1,p,s}}}\right)\times \pmb{d^l_s} \times \epsilon_{t,s}
$$ {#eq-dd3}

@@

$$
  larv_{t,p,s} = ssb_{t-1,p,s} \times\pmb{d^l_s}
$$ {#eq-larvmove}

@@

<!-- obj_p = \Phi \times scale(\sum_f^F \pmb{\phi_{f,p}}) + (1 - \Phi)  scale(\pmb{\sum_s^S\frac{ssb_s}{ssb0_{s,p}}}) -->
